Tidy data structure to support exploration and modeling of temporal-context data

### 2018-05-01 main.Rmd
added opening paragraph in "data semantics"

Temporal-context data consist of observational units indexed at different time points $X_{jt}$, where the $j$\textsuperscript{th} unit takes measurements of $X$ on over time $t$, for $j = 1, \cdots, N$ and $1 \le t \le T$. Time primarily forms the contextual basis of temporal data, but it could arrive in many possible formats. For example, data recorded at fine time resolutions (hours, minutes, and seconds) are typically associated with different time zones and daylight savings. Temporal data often carries with rich information other than the time: multiple observational units of different time lengths, multiple and heterogeneous measured variables, multiple grouping factors involving nested or crossed structures, linking to other data tables, and etc.

In the literature, time series and panel (longitudinal) data are common terms referred to temporal-context data, depending on the research fields. Researchers who are concerned with modelling large $T$ and small $N$ would name as "time series" (serial correlation); those who are interested in modelling small $T$ and large $N$ as "panel data" (asymptotic). The data format is two-dimensional array, but different modelling focuses lend the data input to different representations. A matrix is used to represent multivariate time series where each row represents observations measured at a time point and each column represents a series ("wide" form). This matrix representation requires homogeneity (that is, all the columns must be of same type.), time indices implicitly inferred as attributes or meta-information, series of same time length, and explicit missingness. By contrast, panel data are organised in rectangular form of heterogeneous column types where multiple study subjects are stacked and repeated for its time indices in a single column ("long" form), due to commonly arisen unbalanced panels. This specification requires explicitly declared panel variable and index, which has been implemented in Stata's time series module and R package **plm**. This data organisation appears more flexible than matrix in supporting explicit time index, multiple subjects of different time lengths, and implicit missing values. [R]

Temporal data can often be aggregated in a manner that exhibits a nested or crossed structure, also known as hierarchical or grouped time series [@fpp]. For example, in a manufacturing setting, a company can add up every store’s sales by region, by state and by country, which gives a strictly hierarchical time series; alternatively they can group the sales for each product together based on common attributes such as store, brand, price range and so forth, which leads to a non-hierarchical structure---a grouped time series. The R package **hts** is the implementation of this type of time series. But it is frustrating to create and work with such data objects due to a machine-oriented interface. [R]

@wickham2014tidy formulated the conceptual framework of tidy data: (1) each variable forms a column; (2) each observation forms a row; (3) each type of observational unit forms a table. These principles attempt to standardize the mapping from the semantics of a dataset to its structure and facilitate data analysis in a coherent way. A grammar of data manipulation. [R]

This paper proposes a unified data representation of temporal-context data, blending time series of nested and crossed factors into a two-dimensional column-homogeneous array in a long format. By leveraging the "tidy data" principles, observations and variables position and bridge their meanings in both physical and underlying structures. Data manipulation involves in transforming either observations or variables, or both, which can be described and achieved with a collection of shorthand verbs. A chain of data transformations lend itself to a data pipeline.

@@

A modern re-imagining of time series should provide heterogeneous data types and time indices as explicitly declared data column. This can be achieved using a "data frame" in R or other statistical languages to represent a tabular format, instead of "matrix". Tidy temporal-context data at least consists of:

1. index: an explicitly declared data variable contains time indices, such as date-times, year-months, years and etc.
2. key: a set of grouping factors uniquely identifies each unit that measurements take place on over time, which may include single or multiple columns.
3. interval: data with regular time interval results in a common time interval in one table.

In SQL database, a primary key [@codd_relational_1970], which uniquely defines each record in a database, is equivalent to the composition of "index" and "key".

@@

A pipeline exhibits a hierarchy of data operations: (1) atomic (1-dimensional) vectors (`mean(variale)`) --> (2) (2-dimensional) data table (`summarise()`) --> repeat step (1) and (2) to form a chain.

@@

* **row-wise**: `filter()`, `slice()`, `arrange()`, `fill_na()`
* **column-wise**: `mutate()`, `select()`, `summarise()`, `tsummarise()`
* **rolling window**: `slide()`, `tile()`, `stretch()`
* **statistics**: `lag()`, `diff()`, `acf()`

>>>>

Temporal-context data consist of observational units indexed at different time points $X_{jt}$, where the $j$\textsuperscript{th} unit takes measurements of $X$ on over time $t$, for $j = 1, \dots, N_{i}$ and $1 \le t \le T$. Time primarily forms the contextual basis of temporal data, but it could arrive in many possible formats. For example, data recorded at fine time resolutions (hours, minutes, and seconds) are typically associated with different time zones and daylight savings. Temporal data often carries with rich information other than the time: multiple observational units of different time lengths, multiple and heterogeneous measured variables, multiple grouping factors involving nested or crossed structures, linking to other data tables, and etc.

In the literature, time series and panel (longitudinal) data are common terms referred to temporal-context data, depending on the research fields. Researchers who are concerned with modelling large $T$ and small $N$ would name as "time series" (serial correlation); those who are interested in modelling small $T$ and large $N$ as "panel data" (asymptotic). The data format is two-dimensional array, but different modelling focuses lend the data input to different representations. A matrix is used to represent multivariate time series where each row represents observations measured at a time point and each column represents a series ("wide" form). This matrix representation requires homogeneity (that is, all the columns must be of same type.), time indices implicitly inferred as attributes or meta-information, series of same time length, and explicit missingness. By contrast, panel data are organised in rectangular form of heterogeneous column types where multiple study subjects are stacked and repeated for its time indices in a single column ("long" form), due to commonly arisen unbalanced panels. This specification requires explicitly declared panel variable and index, which has been implemented in Stata's time series module and R package **plm**. Evidently, this data organisation appears more flexible than matrix in supporting explicit time index, multiple subjects of different time lengths, and implicit missing values. [R]

Temporal data can often be aggregated in a manner that exhibits a nested or crossed structure, also known as hierarchical or grouped time series [@fpp]. For example, in a manufacturing setting, a company can add up every store’s sales by region, by state and by country, which gives a strictly hierarchical time series; alternatively they can group the sales for each product together based on common attributes such as store, brand, price range and so forth, which leads to a non-hierarchical structure---a grouped time series. The R package **hts** is the implementation of this type of time series. But it is frustrating to create and work with such data objects due to non-extensible and non-modified. Variational data aggregation (some groups no need aggregation). [R]

@wickham2014tidy coined the term "tidy data" and formalized the processing from messy data to "tidy data". These principles attempt to standardize the mapping from the semantics of a dataset to its structure and facilitate data analysis in a coherent way. Based on the systematic structuring principles, a grammar of data manipulation. [R]

This paper proposes a unified data representation of temporal-context data, blending time series of nested and crossed factors into a two-dimensional column-homogeneous array in a long format. By leveraging the "tidy data" principles, observations and variables position and bridge their meanings in both physical and internal structures. Data manipulation involves in transforming either observations or variables, or both, which can be described and achieved with a collection of shorthand operators. A chain of data transformations lend itself to a data pipeline.

@@

The choice of representation of temporal-context data is made from a data-centric perspective, which is taken in the light of the operations that are to be performed on the data. Firstly, a data set must be structured in a "tidy" layout. Secondly, declaring the data set to contain temporal observations is determined by "index" that represent time and "key" that uniquely identifies each unit that measurements take place on over time. The composition of index and key uniquely defines each observation in a data table, which is equivalent to a primary key [@codd_relational_1970] in a relational database.

@@

Literature about data pipeline.

A pipeline exhibits a hierarchy of data operations: (1) atomic (1-dimensional) vectors (`mean(variable)`) --> (2) (2-dimensional) data table (`summarise()`) --> repeat step (1) and (2) to form a chain. Reversely, a data pipeline is decomposed into rectangular blocks, and then into atomic strips. (UNIX pipe)

@@

* **row-wise**: `filter()`, `slice()`, `arrange()`, `fill_na()`
* **column-wise**: `mutate()`, `select()`, `summarise()`, `tsummarise()`
* **rolling window**: `slide()`, `tile()`, `stretch()`
* **time-wise**: `lag()`, `diff()`, `acf()`
